Semantics for the $ \lambda $-calculus

### 2024-04-15 report/chapters/previous-work.tex
Verify that display maps are well-defined

Now, let $ C $ be a category with a subtype of display maps $ D \subseteq C $. We will denote the subtype of display maps from $ X $ to $ A $ with $ D(X, A) \subseteq C(X, A) $. For any $ A : C $, we get a category $ (C \downarrow_D A) $ as a full subcategory of the slice category $ (C \downarrow A) $, with as objects the display maps $ f: D(X, A) $. A morphism from $ f: D(X, A) $ to $ g: D(Y, A) $ is a morphism $ \varphi: C(X, Y) $ such that the following diagram commutes.
\begin{center}
  \begin{tikzcd}
    X \arrow[rd, "f"'] \arrow[rr, "\varphi"] & & Y \arrow[ld, "g"]\\
    & A &
  \end{tikzcd}
\end{center}
Note that for the terminal object $ I : C $, $ (C \downarrow_D I) $ is still equivalent to $ C $, since every terminal projection is a display map.

Also note that since display maps are closed under pullbacks, the pullback functor $ (C \downarrow A) \to (C \downarrow B) $ for $ \alpha: C(B, A) $ restricts to a pullback functor $ \alpha^*: (C \downarrow_D A) \to (C \downarrow_D B) $.

@@

\TODO Displayed category?

@@

Taylor shows that these display maps are well-defined, in the sense that they satisfy the three properties mentioned above.
\begin{enumerate}
  \item Given a display map $ (X, f): (\R \downarrow_D A) $ and a morphism $ \alpha: \R(B, A) $. Recall that this means that there merely exists a $ Y: \R^A $ and an isomorphism $ X \xrightarrow \sim \sum_A Y $ in $ (\R \downarrow A) $. Therefore, we have mere existence of a pullback $ \sum_B (Y \circ \alpha) $ of $ \sum_A Y $ along $ \alpha $. For all $ C: \R $ with $ \beta: \R(C, B) $ and $ \gamma: \R(C, \sum_A Y) $ that make the square commute, we have the morphism $ \lambda x_1, (\beta x_1, \pi_2 (\gamma x_1)): \R(C, \sum_B (Y \circ \alpha)) $.
    \begin{center}
      \begin{tikzcd}[column sep = .5in]
        C \arrow[rrrd, bend left, "\gamma"] \arrow[dd, bend right=90, "\beta"'] \arrow[d, dashed, "{\lambda x_1, (\beta x_1, \pi_2 (\gamma x_1))}"] \\
        \sum_B (Y \circ \alpha) \arrow[rrr, "{\lambda x_1, (\alpha (\pi_1 x_1), Y (\alpha (\pi_1 x_1)) (\pi_2 x_1))}"] \arrow[d, "p_{(Y \circ \alpha)}"] &&& \sum_A Y \arrow[d, "p_Y"] & X \arrow[l, "g"] \arrow[ld, "f"]\\
        B \arrow[rrr, "\alpha"] &&& A
      \end{tikzcd}
    \end{center}
    By the isomorphism between $ X $ and $ \sum_A Y $, $ \sum_B (Y \circ \alpha) $ is also a pullback of $ X $: by postcomposing with $ g $ and its inverse, we see that morphisms from $ \sum_B (Y \circ A) $ and $ C $ to $ X $ are equivalent to morphisms to $ \sum_A Y $.

    \TODO: Verify that these morphisms are indeed morphisms.
  \item
  \item Let $ B \to I $ be a terminal projection. Consider $ Y = (\lambda x_1, B) : \R^I $. We have
    \[ \sum_I Y = \lambda x_1, ((\lambda x_2, x_2), B (\pi_2 x_1)) = \langle I \circ \pi_1, B \circ \pi_2 \rangle = I \times B, \]
    which is isomorphic to $ B $.
\end{enumerate}

@@

\TODO Relate indexed types and fibrations.

>>>>

Now, let $ C $ be a category with a subtype of display maps $ D \subseteq C $. We will denote the subtype of display maps from $ X $ to $ A $ with $ D(X, A) \subseteq C(X, A) $. For any $ A : C $, we define the category $ (C \downarrow_D A) $ as a full subcategory of the slice category $ (C \downarrow A) $, with as objects the display maps $ f: D(X, A) $. The morphisms between two objects of $ (C \downarrow_D A) $ are still all the morphisms of $ (C \downarrow A) $ (i.e. the morphisms of $ C $ that commute with the display maps).

Note that for the terminal object $ I : C $, $ (C \downarrow_D I) $ is still equivalent to $ C $, since every terminal projection is a display map.

Also note that since the pullback of display maps against any map exist (and are display maps again), we get a pullback functor $ \alpha^*: (C \downarrow_D A) \to (C \downarrow_D B) $. This is the restriction of the pullback functor $ \alpha^*: (C \downarrow A) \to (C \downarrow B) $, if it exists.

@@

@@

Taylor shows that these display maps indeed satisfy the three properties mentioned above. However, in univalent foundations, there are some subtleties in the case of pullbacks:
\begin{enumerate}
  \item Given a display map $ (X, f): (\R \downarrow_D A) $ and a morphism $ \alpha: \R(B, A) $. Recall that this means that there merely exists a $ Y: \R^A $ and an isomorphism $ X \xrightarrow \sim \sum_A Y $ in $ (\R \downarrow A) $. Therefore, we have mere existence of a pullback $ \sum_B (Y \circ \alpha) $ of $ \sum_A Y $ along $ \alpha $. For all $ C: \R $ with $ \beta: \R(C, B) $ and $ \gamma: \R(C, \sum_A Y) $ that make the square commute, we have the morphism $ \lambda x_1, (\beta x_1, \pi_2 (\gamma x_1)): \R(C, \sum_B (Y \circ \alpha)) $.
    \begin{center}
      \begin{tikzcd}[column sep = .5in]
        C \arrow[rrrd, bend left, "\gamma"] \arrow[dd, bend right=90, "\beta"'] \arrow[d, dashed, "{\lambda x_1, (\beta x_1, \pi_2 (\gamma x_1))}"] \\
        \sum_B (Y \circ \alpha) \arrow[rrr, "{\lambda x_1, (\alpha (\pi_1 x_1), Y (\alpha (\pi_1 x_1)) (\pi_2 x_1))}"] \arrow[d, "p_{(Y \circ \alpha)}"] &&& \sum_A Y \arrow[d, "p_Y"] & X \arrow[l, "g"] \arrow[ld, "f"]\\
        B \arrow[rrr, "\alpha"] &&& A
      \end{tikzcd}
    \end{center}
    By the isomorphism between $ X $ and $ \sum_A Y $, $ \sum_B (Y \circ \alpha) $ is also a pullback of $ X $: by postcomposing with $ g $ and its inverse, we see that morphisms from $ \sum_B (Y \circ A) $ and $ C $ to $ X $ are equivalent to morphisms to $ \sum_A Y $.
  \item Given display maps $ f: B \to A $ and $ g: C \to B $. We have $ X : \R^A $ and $ Y: \R^B $ with isomorphisms $ s: B \xrightarrow \sim \sum_A X $ and $ t: C \xrightarrow \sim \sum_B Y $. We need to show that $ f \circ g: C \to A $ is also a display map.
    \begin{center}
      \begin{tikzcd}
        C \arrow[d, "g"] \arrow[r, "\sim", "t"'] & \sum_B Y \arrow[ld, "p_Y"] \arrow[r, "u"', "\sim"] & \sum_A Z \arrow[ldld, bend left, "p_Z"]\\
        B \arrow[d, "f"] \arrow[r, "\sim", "s"'] & \sum_A X \arrow[ld, "p_X"]\\
        A
      \end{tikzcd}
    \end{center}
    Intuitively, $ \sum_B Y \cong \sum_A Z $ represents
    \[ \sum_{b: \sum_{a: A} X a} Y b \cong \sum_{a: A} \sum_{x: X a} Y(a, x), \]
    by the isomorphism that relates $ ((a, x), y) $ to $ (a, (x, y)) $. Therefore, consider
    \[ Z = \lambda x_1 x_2, (X x_1 (\pi_1 x_2), Y (s^{-1} (x_1, \pi_1 x_2)) (\pi_2 x_2)): \R^A. \]
    we have an isomorphism $ u: \sum_B Y \xrightarrow \sim \sum_A Z $ given by:
    \[ u = \lambda x_1, (A a, (X a x, Y b y)) \]
    with
    \[ b = \pi_1 x_1;\quad a = \pi_1 (s b); \quad x = \pi_2 (s b); \quad y = \pi_2 x_1, \]
    and
    \[ u^{-1} = \lambda x_1, (s^{-1}(A a, X a x), Y b y) \]
    with
    \[ a = \pi_1 x_1; \quad x = \pi_1 (\pi_2 x_1); \quad b = s^{-1}(a, x); \quad y = \pi_2 (\pi_2 x_1) \]
    and then $ u \circ t $ is an isomorphism in $ (\R \downarrow A) $ between $ (C, f \circ g) $ and $ (\sum_A Z, p_Z) $.

    % \TODO: Verify that
    % \begin{itemize}
    %   \item $ Z: \R^A $
    %     \begin{align*}
    %       &Z x_1 (Z x_1 x_2)\\
    %       &= (X x_1 (X x_1 (\pi_1 x_2)), Y (s^{-1} (x_1, (X x_1 (\pi_1 x_2)))) (Y (s^{-1} (x_1, (\pi_1 x_2))) (\pi_2 x_2)))\\
    %       &= (X x_1 (\pi_1 x_2), Y (s^{-1} (x_1, \pi_1 x_2)) (\pi_2 x_2))\\
    %       &= Z x_1 x_2.
    %     \end{align*}
    %   \item $ u $ is a morphism
    %     \begin{align*}
    %       \sum_A Z (u (\sum_B Y x_1))
    %       &= \sum_A Z (u (B (\pi_1 x_1), Y (\pi_1 x_1) (\pi_2 x_1)))\\
    %       &= \sum_A Z (\pi_1 (s (\pi_1 x_1)), (\pi_2 (s (\pi_1 x_1)), Y (\pi_1 x_1) (\pi_2 x_1)))\\
    %       &= (A (\pi_1 (s (\pi_1 x_1))), (X (\pi_1 (s (\pi_1 x_1))) ((\pi_2 (s (\pi_1 x_1)))), Y (\pi_1 x_1) (\pi_2 x_1)))\\
    %       &= u x_1.
    %     \end{align*}
    %   \item $ u^{-1} $ is a morphism
    %     \begin{align*}
    %       \sum_B Y (u^{-1} (\sum_A Z x_1))
    %       &= \sum_B Y (u^{-1} (A (\pi_1 x_1), (X (\pi_1 x_1) (\pi_1 (\pi_2 x_1)), Y (s^{-1} ((\pi_1 x_1), \pi_1 (\pi_2 x_1))) (\pi_2 (\pi_2 x_1)))))\\
    %       &= \sum_B Y (s^{-1}(\pi_1 x_1, \pi_1 (\pi_2 x_1)), Y (s^{-1} (\pi_1 x_1, \pi_1 (\pi_2 x_1))) (\pi_2 (\pi_2 x_1)))\\
    %       &= (s^{-1}(\pi_1 x_1, \pi_1 (\pi_2 x_1)), Y (s^{-1} (\pi_1 x_1, \pi_1 (\pi_2 x_1))) (\pi_2 (\pi_2 x_1)))\\
    %       &= u^{-1} x_1
    %     \end{align*}
    %   \item $ u \circ u^{-1} = \id{\sum_A Z} $
    %     \begin{align*}
    %       u(u^{-1} x_1)
    %       &= u(s^{-1} (\pi_1 x_1, \pi_1 (\pi_2 x_1)), Y (s^{-1} (\pi_1 x_1, \pi_1 (\pi_2 x_1))) (\pi_2 (\pi_2 x_1)))\\
    %       &= (A (\pi_1 x_1), (X (\pi_1 x_1) (\pi_1 (\pi_2 x_1)), Y (s^{-1} (\pi_1 x_1, \pi_1 (\pi_2 x_1))) (\pi_2 (\pi_2 x_1))))\\
    %       &= \sum_A Z x_1
    %     \end{align*}
    %   \item $ u^{-1} \circ u = \id{\sum_B Y} $.
    %     \begin{align*}
    %       u^{-1} (u x_1)
    %       &= u^{-1}(\pi_1 (s (\pi_1 x_1)), (\pi_2 (s (\pi_1 x_1)), Y (\pi_1 x_1) (\pi_2 x_1)))\\
    %       &= (B (\pi_1 x_1), Y (\pi_1 x_1) (\pi_2 x_1))
    %       &= \sum_B Y x_1.
    %     \end{align*}
    %   \item $ p_Z \circ u = g \circ p_Y $.
    %     \begin{align*}
    %       p_Z (u x_1)
    %       &= p_Z (\pi_1 (s (\pi_1 x_1)), (\pi_2 (s (\pi_1 x_1)), Y (\pi_1 x_1) (\pi_2 x_1)))\\
    %       &= A (\pi_1 (s (\pi_1 x_1)))\\
    %       &= A (\pi_1 (s (B (\pi_1 x_1))))\\
    %       &= p_X (s (p_Y x_1))\\
    %       g (p_Y x_1).
    %     \end{align*}
    % \end{itemize}
  \item Let $ B \to I $ be a terminal projection. Consider $ Y = (\lambda x_1, B) : \R^I $. We have
    \[ \sum_I Y = \lambda x_1, ((\lambda x_2, x_2), B (\pi_2 x_1)) = \langle I \circ \pi_1, B \circ \pi_2 \rangle = I \times B, \]
    which is isomorphic to $ B $.
\end{enumerate}
\begin{remark}
  There is an awful lot of properties to verify here (e.g. that some terms are elements of some $ \R^A $, that others are morphisms in some $ \R(\sum_A X, \sum_B Y) $, that some diagrams commute), but most of it boils down to writing down the equations that should hold, reducing them via $ \beta $-equality, and then using the facts that objects $ A: \R $ are idempotent, that morphisms $ f: \R(A, B) $ equal $ B \circ f \circ A $ and that objects $ X: \R^A $ satisfy $ X \circ A = X $ and $ X a (X a x) = X a x $ for all $ a $ and $ x $. Note that for a morphism $ f: \sum_A X \to B $,
  \[ f(A x_1, x_2) = f(x_1, x_2) = f(x_1, X x_1 x_2), \]
  so $ f $ `absorbs' such instances of $ A $ and $ X $, so to speak.
\end{remark}

@@
